A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-05-11 paper/paper1/ms.tex
Uncomitted changes

\subsection{Gaia DR3}\label{analysisgaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere 0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint.

@@

\subsubsection{Revised exoplanet orbital parameters}
We assembled a composite lightcurve by stitching TESS Sectors 24, 26, 51, 52, 53, which were reduced with the default SPOC pipeline \citep{2020RNAAS...4..201C}, and lightly post-processed with \texttt{lightkurve} \citep{geert_barentsen_2019_2565212}.  These TESS data exhibit an RMS scatter of better than 1 part-per-thousand (ppt) at native 2-minute sampling.  We fit a Keplerian orbit model to the TESS lightcurve using the \texttt{exoplanet} framework \citep{exoplanet:joss}.  We obtained a best fit orbital period of $4.81011$ days, non-zero eccentricity of 0.19, an impact parameter of 0.46, planet-to-star radius ratio of 0.0823, transit depth of 0.74\%, and $T_0$ of 1958.07933 in BTJD.  These properties are consistent with the previously reported values from \citet{2017AJ....153..211Z} and the updated ephemeris of \citet{2022ApJS..259...62I}.  Figure \ref{fig:TESSoverview} shows an overview of all the TESS Sectors demarcated with the epochs of HPF measurements.  Figure \ref{fig:transit} shows the best-fit orbit overlaid on the detrended TESS lightcurve.  Table \ref{tabOrbit} lists one of the previous orbit determinations and the orbital solution reported here.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation on timescales comparable to the exoplanet orbit. TESS Sectors 51-53 do not show as conspicuous a modulation signal.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.

If we attribute this TESS modulation signal to stellar activity, we obtain a rotation period $P_\mathrm{rot}$ between 4.7 and 5.9 days.  We can constrain the expected rotation period based on our revised stellar radius estimate, stellar rotation rate from $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range may be typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears consistent with only an eccentric orbit model combined with the Geneva isochrone \citep{2017AJ....153..211Z}.

\subsubsection{Lightcurve modulation amplitude}
We computed the peak-to-valley modulation for the 1000-source \emph{TESS} comparison sample, defined as the difference between the 90$^{th}$ to 10$^{th}$ percentile of each normalized \emph{TESS} lightcurve.  Most lightcurves appear consistent with negligible astrophysical variability and/or variability predominated by instrumental artifacts. \object{HAT-P-67}'s 0.36$\%$ lightcurve modulation ranked in the 87$^{th}$ percentile of the ensemble, about 0.5$\times$ above the median.  Sources more variable than HAT-P-67 exhibited 3-5$\times$ its amplitude of modulation.

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
As discussed in Section \ref{analysisgaiaDR3}, the slightly increased distance and radius imply a slightly more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;][]{2016ApJS..222....8D,2016ApJ...823..102C} for HAT-P-67.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation to \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.  We therefore adopt a MIST-based 2$\pm$0.2 Gyr age for the system, in-between the previous Geneva and Dartmouth-based scenarios.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$, slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  Notably, our fit does not jointly model the Doppler tomographic signals reported in \citep{2017AJ....153..211Z}, which would further constrain the impact parameter.  Additionally, our input $\rstar$ was slightly lower than the best-fit value derived from \emph{Gaia DR3} and SED fitting.
%\authorcomment1{Quang Tran: please verify and revise.}

>>>>

\subsection{Gaia DR3}\label{analysisgaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) and isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$ from the measured transit depth, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere $<$0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint.  The luminosity increases by 18\%, to about 10.3 $L_\odot$.

@@

\subsubsection{Revised exoplanet orbital parameters}
We assembled a composite lightcurve by stitching TESS Sectors 24, 26, 51, 52, 53, which were reduced with the default SPOC pipeline \citep{2020RNAAS...4..201C}, and lightly post-processed with \texttt{lightkurve} \citep{geert_barentsen_2019_2565212}.  These TESS data exhibit an RMS scatter of better than 1 part-per-thousand (ppt) at native 2-minute sampling.  We fit a Keplerian orbit model to the TESS lightcurve using the \texttt{exoplanet} framework \citep{exoplanet:joss}.  We obtained revised orbital properties shown in Table \ref{tabOrbit}.  These properties are broadly consistent with the previously reported values from \citet{2017AJ....153..211Z} and the updated ephemeris of \citet{2022ApJS..259...62I}.  Figure \ref{fig:TESSoverview} shows an overview of all the TESS Sectors with a green GP trendline in green highlighting the out-of-transit modulations, and the exoplanet transit model in orange.  Figure \ref{fig:transit} shows the best-fit orbit overlaid in purple on the detrended TESS lightcurve, which is binned in the orange dots.  Table \ref{tabOrbit} lists only one of the previous orbit determinations---also assuming a circular orbit---compared with the orbital solution reported here.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation. TESS Sectors 51-53 do not show as conspicuous a modulation signal, but still show some ostensibly stellar variability.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range may be typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

\subsubsection{Lightcurve modulation amplitude}
\authorcomment1{Slated for removal:}
\textcolor{lightgray}{We computed the peak-to-valley modulation for the 1000-source \emph{TESS} comparison sample, defined as the difference between the 90$^{th}$ to 10$^{th}$ percentile of each normalized \emph{TESS} lightcurve.  Most lightcurves appear consistent with negligible astrophysical variability and/or variability predominated by instrumental artifacts. \object{HAT-P-67}'s 0.36$\%$ lightcurve modulation ranked in the 87$^{th}$ percentile of the ensemble, about 0.5$\times$ above the median.  Sources more variable than HAT-P-67 exhibited 3-5$\times$ its amplitude of modulation.}

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
The slightly increased distance and radius imply an 18\% more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;][]{2016ApJS..222....8D,2016ApJ...823..102C} for HAT-P-67.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation over the lifetime of \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.  We therefore adopt a MIST-based 2$\pm$0.2 Gyr age for the system, in-between the previous Geneva and Dartmouth-based scenarios \citep{2017AJ....153..211Z}.  Additional metallicity and rotation effects could slightly alter the evolutionary tracks and therefore increase uncertainty in inferred ages and masses.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \kms---assuming a circular orbit---slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  Notably, our fit does not jointly model the Doppler tomographic signals reported in \citep{2017AJ....153..211Z}, which could further constrain the impact parameter.  Additionally, our input $\rstar$ was slightly lower than the best-fit value derived from \emph{Gaia DR3} and SED fitting.
\authorcomment1{Quang Tran: please verify and revise.}
